Inference of proto-neutron star properties from gravitational-wave data in core-collapse supernovae

### 2020-11-09 ccsn.tex
Pablo s comemnts answer

\begin{abstract}
The eventual detection of gravitational waves from core-collapse supernovae (CCSN) will help improve our current understanding of the explosion mechanism of massive stars. The stochastic nature of the late post-bounce gravitational wave signal due to \pcd{\sout{matter effects} the non-linear dynamics of the matter involved} and the large number of degrees of freedom of the phenomenon make the source parameter inference problem very challenging. In this paper we take a step towards that goal and present a parameter estimation approach which is based on the gravitational waves associated with \pcd{\sout{convective}} oscillations of proto-neutron stars (PNS)
\pcd{[PCD: the oscillations are driven by convection and other instabilities but I would not call them convective oscillations.]}. Numerical simulations of CCNS have shown that buoyancy-driven g-modes are responsible for a significant fraction of the gravitational wave signal and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations, as demonstrated in~\cite{Torres:2019b}. We use a set of 1D CCSN simulations to build a model that relates the  evolution of the PNS properties with the frequency of the dominant g-mode, which is extracted from the gravitational-wave data using a new algorithm we have developed for our study. The model is used to infer the time evolution of combinations \pcd{[PCD: "of a combination"? (there is only one)]} of the mass and the radius of the PNS. The performance of the method is estimated employing simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state. Considering signals embedded in Gaussian gravitational wave detector noise, we show that it is possible to infer PNS properties for a galactic source using Advanced LIGO and Advanced Virgo data at design sensitivities. Third generation detectors such as Einstein Telescope and Cosmic Explorer will allow to test distances of ${\cal O}(100\, {\rm kpc})$. %\tf{Caveats? Gaussian noise?}
% 
%  Core-collapse supernovae are very important phenomena in the Universe and their expected gravitational
%  waves emission is a promising tool to study the onset of the supernova explosion mechanism. The complexity
%  of the gravitational wave signal due to matter effects and the large degrees of freedom of the phenomena
%  makes the source parameter inference problem very challenging. In this paper, we have considered the
%  proto neutron star oscillation modes that carry out a large fraction of the gravitational wave signal and
%  whose frequency time evolution is linked to the system properties through universal relations as
%  demonstrated in ~\cite{Torres:2019b}. We have developed a simple algorithm to extract from the
%  gravitational wave data the frequency evolution of the gravity mode. A set of 1D core collapse
%  supernova simulations is used to build a model of evolution of the proto neutron star properties
%  with the gravity mode frequency. The model is then used to infer the time evolution of a combinaison of
%  the radius and mass of the proto neutron star. We have then estimated the performance of the method by
%  performing simulations with 2D core collapse supernova waveforms covering a progenitor mass range between
%  11 and 40 solar mass and different equations of state. We have shown that for Advanced LIGO and Advanced
%  Virgo detectors at design sensitivity it will be possible to infer proton neutron star properties for
%  a galactic source, while third generation detectors Einstein Telescope and Cosmic Explorer will allow
%  to test distances of several hundreds of kiloparsec.

@@

\bigskip\noindent\textit{Acknowledgments.--} Work supported by the Spanish Agencia Estatal de Investigaci\'on (PGC2018-095984-B-I00), by the Generalitat Valenciana (PROMETEO/2019/071) and by the European Union’s Horizon 2020 research and innovation (RISE) programme H2020-MSCA-RISE-2017 GrantNo. FunFiCO-777740.
N.C. acknowledges support from National Science Foundation grant PHY-1806990.
R.M. gratefully acknowledges support by the James Cook Fellowship from Government funding, administered by the Royal Society Te  Ap\={a}rangi. R.M. and P.M. also acknowledge funding from  DFG Grant KI 1443/3-2 and thank the Centre for eResearch at the University of Auckland for their technical support.
P.C.-D. acknowledges the support of the Spanish {\it Ramon y Cajal} programme (RYC-2015-19074) supporting his research.
\begin{appendices}
\input{appendixA}
\end{appendices}
%\bibliographystyle{unsrt}
\bibliography{biblio}

>>>>

\begin{abstract}
The eventual detection of gravitational waves from core-collapse supernovae (CCSN) will help improve our current understanding of the explosion mechanism of massive stars. The stochastic nature of the late post-bounce gravitational wave signal due to the non-linear dynamics of the matter involved and the large number of degrees of freedom of the phenomenon make the source parameter inference problem very challenging. In this paper we take a step towards that goal and present a parameter estimation approach which is based on the gravitational waves associated with oscillations of proto-neutron stars (PNS)
Numerical simulations of CCNS have shown that buoyancy-driven g-modes are responsible for a significant fraction of the gravitational wave signal and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations, as demonstrated in~\cite{Torres:2019b}. We use a set of 1D CCSN simulations to build a model that relates the  evolution of the PNS properties with the frequency of the dominant g-mode, which is extracted from the gravitational-wave data using a new algorithm we have developed for our study. The model is used to infer the time evolution of a combination of the mass and the radius of the PNS. The performance of the method is estimated employing simulations of 2D CCSN waveforms covering a progenitor mass range between 11 and 40 solar masses and different equations of state. Considering signals embedded in Gaussian gravitational wave detector noise, we show that it is possible to infer PNS properties for a galactic source using Advanced LIGO and Advanced Virgo data at design sensitivities. Third generation detectors such as Einstein Telescope and Cosmic Explorer will allow to test distances of ${\cal O}(100\, {\rm kpc})$. %\tf{Caveats? Gaussian noise?}
% 
%  Core-collapse supernovae are very important phenomena in the Universe and their expected gravitational
%  waves emission is a promising tool to study the onset of the supernova explosion mechanism. The complexity
%  of the gravitational wave signal due to matter effects and the large degrees of freedom of the phenomena
%  makes the source parameter inference problem very challenging. In this paper, we have considered the
%  proto neutron star oscillation modes that carry out a large fraction of the gravitational wave signal and
%  whose frequency time evolution is linked to the system properties through universal relations as
%  demonstrated in ~\cite{Torres:2019b}. We have developed a simple algorithm to extract from the
%  gravitational wave data the frequency evolution of the gravity mode. A set of 1D core collapse
%  supernova simulations is used to build a model of evolution of the proto neutron star properties
%  with the gravity mode frequency. The model is then used to infer the time evolution of a combinaison of
%  the radius and mass of the proto neutron star. We have then estimated the performance of the method by
%  performing simulations with 2D core collapse supernova waveforms covering a progenitor mass range between
%  11 and 40 solar mass and different equations of state. We have shown that for Advanced LIGO and Advanced
%  Virgo detectors at design sensitivity it will be possible to infer proton neutron star properties for
%  a galactic source, while third generation detectors Einstein Telescope and Cosmic Explorer will allow
%  to test distances of several hundreds of kiloparsec.

@@

\bigskip\noindent\textit{Acknowledgments.--} Work supported by the Spanish Agencia Estatal de Investigaci\'on (PGC2018-095984-B-I00), by the Generalitat Valenciana (PROMETEO/2019/071) and by the European Union’s Horizon 2020 research and innovation (RISE) programme H2020-MSCA-RISE-2017 GrantNo. FunFiCO-777740.
N.C. acknowledges support from National Science Foundation grant PHY-1806990.
R.M. gratefully acknowledges support by the James Cook Fellowship from Government funding, administered by the Royal Society Te  Ap\={a}rangi. R.M. and P.M.-R. also acknowledge funding from  DFG Grant KI 1443/3-2 and thank the Centre for eResearch at the University of Auckland for their technical support.
P.C.-D. acknowledges the support of the Spanish {\it Ramon y Cajal} programme (RYC-2015-19074) supporting his research.
\begin{appendices}
\input{appendixA}
\end{appendices}
%\bibliographystyle{unsrt}
\bibliography{biblio}

### 2020-11-09 conclusion.tex
Pablo s comemnts answer

% !TEX root = ccsn.\TeX
\section{Conclusions}
\label{sec:conclusion}

The detection of GW from CCSN may help improve our current understanding of the explosion mechanism of massive stars.  
In this paper we have provided estimates of the detectability prospects of CCSN signals using an approach based on GW associated with convective oscillations of PNS. As shown by~\cite{Torres:2019b} buoyancy-driven g-modes are excited in numerical simulations of CCSN and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations. Such modes are responsible for a significant fraction of the highly stochastic GW emitted after core bounce. The findings reported in this paper suggest that PNS asteroseismology might be within reach of current and third-generation GW detectors.

>>>>

% !TEX root = ccsn.tex
\section{Conclusions}
\label{sec:conclusion}

The detection of GW from CCSN may help improve our current understanding of the explosion mechanism of massive stars.  
In this paper we have proposed an explorary method to infer PNS properties using an approach based on GW associated with convective oscillations of PNS. As shown by~\cite{Torres:2019b} buoyancy-driven g-modes are excited in numerical simulations of CCSN and their time-frequency evolution is linked to the physical properties of the compact remnant through universal relations. Such modes are responsible for a significant fraction of the highly stochastic GW emitted after core bounce. The findings reported in this paper suggest that PNS asteroseismology might be within reach of current and third-generation GW detectors.

### 2020-11-09 introduction.tex
Pablo s comemnts answer

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be 
%are one of the main candidates for 
the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors. The emitted GWs could be detected with the advanced ground-based GW detector network, Advanced LIGO (aLIGO)~\citep{TheLIGOScientific:2014jea}, Advanced Virgo (AdV)~\citep{TheVirgo:2014hva} and
KAGRA~\citep{Aso:2013eba}, within $\sim 5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within $50$~kpc and, for some extreme models, up to $5-30$ Mpc \citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated with long GRBs 
\citep{Chapman:2007}) making this possibility a subdominant channel of detection with an event rate of $\sim 10^{-4} \rm{yr}^{-1}$\mab{[add ref?]}. For the results discussed in this work we only consider neutrino-driven CCSN. \tf{This last sentence must be mentioned somewhere (or maybe not) but perhaps here is not the best place.}\mab{MAB:I'm in favor of deleting it as it is rather obvious we are considering neutrino driven CCSN} Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

@@

Using results from 2D CCSN numerical simulations as a background with respect to which to solve an eigenvalue problem~\citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum of the region within the shock (including the PNS and the post-shock region) 
shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}). \tf{This last sentence might be improved.} \mab{MAB: What about: Computing the mode spectrum of the region within the shock (including the PNS and the post-shock region) consists in solving an eigenvalue problem using background simulations. Using results from 2D CCSN numerical simulations \citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}). }

>>>>

The life of sufficiently massive stars, i.e.~those born with masses between $\sim 8$~M$_\odot$ and $\sim 120$~M$_\odot$, ends with the collapse of {the} iron core under {its} own gravity, leading {to} the formation of a neutron star {(NS)} or a black hole (BH), {and} followed (typically but not necessarily in the BH case) by {a supernova} explosion. Nearby core-collapse supernova (CCSN) explosions are expected to be sources of gravitational waves (GWs) and they could be 
%are one of the main candidates for 
the next great discovery of current ground-based observatories. However, these are relative rare events. A neutrino-driven explosion \citep{Bethe:1990} is the most likely outcome in the case of slowly rotating cores, which are present in the bulk of CCSN progenitors. The emitted GWs could be detected with the advanced ground-based GW detector network, Advanced LIGO (aLIGO)~\citep{TheLIGOScientific:2014jea}, Advanced Virgo (AdV)~\citep{TheVirgo:2014hva} and
KAGRA~\citep{Aso:2013eba}, within $\sim 5$~kpc \citep{Gossan:2016,TargetedSNSearchO12}. Such a galactic event has a rate of about $2-3$ per century \citep{Adams:2013,Rozwadowska:2021}.
For the case of rapidly rotating progenitor cores the result is likely a magneto-rotational explosion, yielding  a more powerful GW signal that could be detected within $50$~kpc and, for some extreme models, up to $5-30$ Mpc \citep{Gossan:2016,TargetedSNSearchO12}. However, only about $1\%$ of the electromagnetically observed events show signatures of fast rotation (broad-lined type Ic SNe \citep{Li:2011b} or events associated with long GRBs \citep{Chapman:2007}) making this possibility a subdominant channel of detection with a galactic event rate of $\sim 10^{-4} \rm{yr}^{-1}$.
%For the results discussed in this work we only consider neutrino-driven CCSN of non-rotating stellar cores.%\tf{This last sentence must be mentioned somewhere (or maybe not) but perhaps here is not the best place.}\mab{MAB:I'm in favor of deleting it as it is rather obvious we are considering neutrino driven CCSN}
Despite the low rates, CCSN are of great scientific interest because they produce complex GW signals which could provide significant clues about the physical processes at work after the gravitational collapse of stellar cores.

@@

Using results from 2D CCSN numerical simulations as a background~\citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum of the region within the shock (including the PNS and the post-shock region) 
shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}).
%\tf{This last sentence might be improved.} \mab{MAB: What about: Computing the mode spectrum of the region within the shock (including the PNS and the post-shock region) consists in solving an eigenvalue problem using background simulations. Using results from 2D CCSN numerical simulations \citep{Torres:2018, Torres:2019a} found that the eigenmode spectrum shows a good match to the mode frequencies and to the features observed in the GW spectrum of the same simulations (specially when space-time perturbations are included \citep{Torres:2019a}). } \pcd{[PCD: I am not convinced about the first sentence of mab's proposal. Maybe if we just delete "with respect to which to solve an eigenvalue problem" the sentence becomes more readable]}

### 2020-11-09 results.tex
Pablo s comemnts answer

To better quantify how well we reconstruct the ratio we also consider  the mean of the relative error of $r_i$ along 
the track of the spectrogram, $\Delta$,  
\begin{equation}
\Delta=\frac{1}{N}\sum_1^N\frac{|r_i-r_i^0|}{r_i^0}\,.
\end{equation}
The values of $\Delta$ for each of the 100 noise realizations are shown as a function of the distance
in the bottom panel of Fig.~\ref{fig:s20results}. For a source located up to $\sim$\unit[9]{kpc} the relative error
remains smaller than 20\%. At closer distances $\Delta$ is smaller but it does not vanish. This reflects the approximate nature of the model used for $r$. It is nevertheless remarkable that, on average, one can reconstruct the ratio time series with a good
precision up to distances of $\sim$ 9 kpc (for this particular waveform) with a coverage value
larger than 80\%. We note that there are a few noise realizations for distances below 9 kpc for which
$\Delta$ takes large values, indicating that the method fails to accurately reconstruct the ratio in those cases. \mab{The main cause of failure is the split of the ridge in few blobs when the signal becomes weaker and weaker.}
%\tf{Do we know why that is the case? What is intrinsically different from one noise realisation to another that makes the method more prone to failure? I think we should provide some explanation here, if we have it.}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_coverage_allwvfs}
 \caption{Median of the coverage for the eight CCSN waveforms of the {\it test set} embedded in aLIGO noise as a function of the  distance to the source\pcd{[PCD: if the solid line is the median, what are the symbols? Same for figures 6 and 7.]}. \mab{The ``no signal'' line shows the median of coverage in absence of any signal. In this example, the median is null and overlaps with the horizontal axis. The blue band boundaries are given by the $\mathrm{5^{th}}$ and $\mathrm{95^{th}}$ percentiles of coverage in absence of any signal.}} \label{fig:aLIGO_cov_allwvf}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_delta_allwvfs}
 \caption{Same as Fig.~\ref{fig:aLIGO_cov_allwvf} but for the relative error $\Delta$.}
% Median of $\Delta$ for 8 CCSN waveforms embedded in aLIGO noise and located at different distance from the Earth. The ``no signal'' line and band show the median and first and third quartile of $\Delta$ in absence of any signal.} 
\label{fig:aLIGO_prec_allwvf}
\end{figure}

We have tested that the method does not depend on the specific features of the waveform of model {\texttt s20S} by repeating the procedure for the remaining seven waveforms of the {\it test set} described in Section \ref{sec:simulations} covering
a large range of progenitor masses. Figure \ref{fig:aLIGO_cov_allwvf} shows that apart from model {\texttt s11} and to a lesser extent model {\texttt s20S}, the ratio is well reconstructed for all waveforms up to a distance of $\sim$ 15kpc. In an
effort to better determine the maximal distance of the source at which we can reconstruct the ratio we have run 100 simulations without injecting a signal and have measured the corresponding coverage for the reconstructed ratios.
\mab{The median of the coverage as well as the band defined by the $\mathrm{5^{th}}$ and $\mathrm{95^{th}}$ percentiles are shown in Figure \ref{fig:aLIGO_cov_allwvf}.}
The noise only median value is identically zero in this case. However, note that it could be different
from zero because
the g-mode reconstruction algorithm is looking for a continuously  increasing frequency track
in the spectrogram, starting between 0 and 200 Hz, where we expect the GW signal to be.
This is enhancing the probability of overlap. This effect explains why certain values of overlap can reach
values as high as 80\% even when no signal is added to the noise. \tf{I don't understand this last comment.} \mab{Toni: is that clearer?}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_all3G}
  \caption{Median of the relative error $\Delta$ as a function of the  distance to the source for the CCSN waveform model {\texttt s20S} embedded in third-generation detector noise. \pcd{[PCD: Should we show the corresponding
  blue ranges for the pure noise case?]}}
  \label{fig:s20--SFHo_all3G}
\end{figure}

>>>>

To better quantify how well we reconstruct the ratio we also consider  the mean of the relative error of $r_i$ along 
the track of the spectrogram, $\Delta$,  
\begin{equation}
\Delta=\frac{1}{N}\sum_1^N\frac{|r_i-r_i^0|}{r_i^0}\,.
\end{equation}
The values of $\Delta$ for each of the 100 noise realizations are shown as a function of the distance
in the bottom panel of Fig.~\ref{fig:s20results}. For a source located up to $\sim$\unit[9]{kpc} the relative error
remains smaller than 20\%. At closer distances $\Delta$ is smaller but it does not vanish. This reflects the approximate nature of the model used for $r$. It is nevertheless remarkable that, on average, one can reconstruct the ratio time series with a good
precision up to distances of $\sim$ 9 kpc (for this particular waveform) with a coverage value
larger than 80\%. We note that there are a few noise realizations for distances below 9 kpc for which
$\Delta$ takes large values, indicating that the method fails to accurately reconstruct the ratio in those cases. The main cause of failure is the split of the ridge in few blobs when the signal becomes weaker and weaker.
%\tf{Do we know why that is the case? What is intrinsically different from one noise realisation to another that makes the method more prone to failure? I think we should provide some explanation here, if we have it.}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_coverage_allwvfs}
 \caption{Data points (``+'') show the median of the coverage for the eight CCSN waveforms of the {\it test set} embedded in aLIGO noise as a function of the  distance to the source. The solid lines are smoothing splines. The ``no signal'' line shows the median of coverage in absence of any signal. In this example, the median is null and overlaps with the horizontal axis. The blue band boundaries are given by the $\mathrm{5^{th}}$ and $\mathrm{95^{th}}$ percentiles of coverage in absence of any signal.} \label{fig:aLIGO_cov_allwvf}
\end{figure}

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/aLIGO_delta_allwvfs}
  \caption{Data points (``+'') show the median of the relative error $\Delta$ for the eight CCSN waveforms of the {\it test set} embedded in aLIGO noise as a function of the distance to the source. The solid lines are smoothing splines. The ``no signal'' line shows the median of $\Delta$ in absence of any signal. The blue band boundaries are given by the $\mathrm{5^{th}}$ and $\mathrm{95^{th}}$ percentiles of $\Delta$ in absence of any signal.}\label{fig:aLIGO_prec_allwvf}
\end{figure}

We have tested that the method does not depend on the specific features of the waveform of model {\texttt s20S} by repeating the procedure for the remaining seven waveforms of the {\it test set} described in Section \ref{sec:simulations} covering
a large range of progenitor masses. Figure \ref{fig:aLIGO_cov_allwvf} shows that apart from model {\texttt s11} and to a lesser extent model {\texttt s20S}, the ratio is well reconstructed for all waveforms up to a distance of $\sim$ 15kpc. In an
effort to better determine the maximal distance of the source at which we can reconstruct the ratio we have run 100 simulations without injecting a signal and have measured the corresponding coverage for the reconstructed ratios.
The median of the coverage as well as the band defined by the $\mathrm{5^{th}}$ and $\mathrm{95^{th}}$ percentiles are shown in Figure \ref{fig:aLIGO_cov_allwvf}.
The noise only median value is identically zero in this case. However, note that it could be different
from zero because
the g-mode reconstruction algorithm is looking for a continuously  increasing frequency track
in the spectrogram, starting between 0 and 200 Hz, where we expect the GW signal to be.
This is enhancing the probability of overlap. This effect explains why certain values of overlap can reach
values as high as 80\% even when no signal is added to the noise. \tf{I don't understand this last comment.} \mab{Toni: is that clearer?}

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.5\textwidth]{plots/s20--SFHo_all3G}
  \caption{Median of the relative error $\Delta$ as a function of the  distance to the source for the CCSN waveform model {\texttt s20S} embedded in third-generation detector noise. \pcd{[PCD: Should we show the corresponding blue ranges for the pure noise case?]}\mab{MAB: We would need to show 5 noise ranges. I haven't tried what it looks like. Could give a try}}
  \label{fig:s20--SFHo_all3G}
\end{figure}
